---
title: "Phyloseq"
author: "Jakub Kreisinger"
date: "1/18/2024"
output: html_document
---

# Packages and custom functions

```{r}
library(phyloseq)
library(ggplot2)
library(vegan)
library(ShortRead)
library(ape)
library(dada2)

# this function removes ASVs that were not detected in both technical replicates
dupl.concensus<-function(PHYLOS,NAMES){
  
  # exclude nonduplicated samples
  IDS<-as.character(data.frame(sample_data(PHYLOS))[,NAMES])
  IDS.dupl<-IDS[duplicated(IDS)]
  
  PHYLOSEQ<-prune_samples(IDS%in%IDS.dupl, PHYLOS)
  if(length(IDS.dupl)*2<length(IDS)) {NONUPLICATED<-prune_samples(!IDS%in%IDS.dupl, PHYLOS)
  print(paste("Following names are nonduplicated",sample_names(NONUPLICATED)))}
  
  CATS<-as.character(data.frame(sample_data(PHYLOSEQ))[,NAMES])
  CATS2<-levels(factor(CATS))
  OTU_TAB<-otu_table(PHYLOSEQ)
  rownames(OTU_TAB)<-CATS
  
  # i<-5
  for (i in 1:length(CATS2))
  {
    # print(CATS2[i])
    FILTER.act<-colSums(OTU_TAB[rownames(OTU_TAB)==CATS2[i],]>0)>1
    OTU_TAB[rownames(OTU_TAB)==CATS2[i],]
    OTU_TAB[rownames(OTU_TAB)==CATS2[i],]<-t(apply(OTU_TAB[rownames(OTU_TAB)==CATS2[i],],1,function(x) x*FILTER.act))
  }
  
  rownames(OTU_TAB)<-sample_names(PHYLOSEQ)
  otu_table(PHYLOSEQ)<-OTU_TAB
  PHYLOSEQ.clean<-prune_taxa(taxa_sums(PHYLOSEQ)>0,PHYLOSEQ)
  
  PHYLOSEQ.clean
}


# this function merge technical replicates in phyloseq object
merge.duplicates<-function(PHYLOSEQ,NAMES){
  CATS<-as.character(data.frame(sample_data(PHYLOSEQ))[,NAMES])
  sample_data(PHYLOSEQ)$duplic.id<-CATS
  SAMDAT<-sample_data(PHYLOSEQ)
  SAMDAT.sub<-subset(SAMDAT,duplicated(CATS)==F)
  FASTA<-refseq(PHYLOSEQ)
  rownames(SAMDAT.sub)<-SAMDAT.sub$duplic.id
  PHYLOSEQ.merge<-merge_samples(PHYLOSEQ,"duplic.id")
  sample_data(PHYLOSEQ.merge)<-SAMDAT.sub
  PHYLOSEQ.merge<-merge_phyloseq(PHYLOSEQ.merge,FASTA)
  PHYLOSEQ.merge
}


```


# ASV abundance matrix

```{r}
load("/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/otutab_16s_pseudo.R")
OTU_T<-otu_table(seqtab,taxa_are_rows = F)
```



# Metadata for phyloseq

```{r}
META<-read.delim("/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/Sample_data_tabulky_areny/Sample_data_tabulka_areny1816cele.txt")

NAMES<-paste0(META$adaptor_R_seq,"_",META$adaptor_F_seq,"_-F_",META$primer_F,"R_trus_",META$RUN)
sum(!NAMES%in%sample_names(OTU_T))

META<-sample_data(META)
sample_names(META)<-NAMES

```

# Phyloseq

Phyloseq including abundance matrix metadata and reference sequences

```{r pressure, echo=FALSE}
REFSEQ<-DNAStringSet(colnames(OTU_T))
names(REFSEQ)<-colnames(OTU_T)
PHLS<-merge_phyloseq(OTU_T,META,REFSEQ)
dim(OTU_T)
dim(META)

PHLS<-prune_taxa(taxa_sums(PHLS)>0,PHLS)
PHLS

```

# Chimeric sequences

Chumeras were detected using *usearch11 -uchime2_ref*

```{r}
SEQS<-refseq(PHLS)
writeFasta(SEQS,file = "/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/haplo.fasta")

# Unix command
 ~/software/bin/usearch11.0.667_i86linux32 -uchime2_ref haplo.fasta -db ~/DB/DADA2/silva_nr99_v138_train_set.fa   -uchimeout out.txt -strand plus -mode balanced -nonchimeras haplo.uchime.fasta


CHIM<-read.delim("/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/out.txt",header=F)
head(CHIM)
dim(CHIM)
summary(as.factor(CHIM$V3))

NONCHIM<-CHIM$V1[CHIM$V3!="Y"]

FILT<-names(SEQS)%in%NONCHIM

PHLS<-prune_taxa(FILT,PHLS)
save(PHLS,file = "/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/Phyloseq_duplicates.R")
```

# Technical duplicates

Technical duplicates are identified and saved in separate databases (dupl.1 and dupl.2)

```{r}
PHLS<-prune_samples(sample_sums(PHLS)>600,PHLS)

KEEP<-sample_data(PHLS)$ID_D[duplicated(sample_data(PHLS)$ID_D)]
PHYLOSEQ.DUPL<-prune_samples(sample_data(PHLS)$ID_D%in%KEEP,PHLS)
PHYLOSEQ.NONDUPL<-prune_samples(!sample_data(PHLS)$ID_D%in%KEEP,PHLS)

DUPL<-duplicated(sample_data(PHYLOSEQ.DUPL)$ID_D)

dupl.1<-prune_samples(DUPL==T,PHYLOSEQ.DUPL)
dupl.2<-prune_samples(DUPL==F,PHYLOSEQ.DUPL)

```

# Alpha diversity correlation

This chunk checks whether the alpha diversity (Shannon entropy) of the community is consistent in the technical replicates.

```{r}

DIV1<-estimate_richness(dupl.1, measures = "Shannon")
DIV2<-estimate_richness(dupl.2, measures = "Shannon")

DIV1<-data.frame(DIV1,sample_data(dupl.1))
DIV2<-data.frame(DIV2,sample_data(dupl.2))
DIV1$seqsums1<-sample_sums(dupl.1)
DIV2$seqsums2<-sample_sums(dupl.2)

DIV2<-DIV2[match(DIV1$ID_D, DIV2$ID_D),]

DF<-data.frame(SH1=DIV1$Shannon,
               SH2=DIV2$Shannon,
               NAME=DIV1$ID_D,
               seqsums1=DIV1$seqsums1,
               seqsums2=DIV2$seqsums2,
               sample=DIV1$ID_D)


DF$OUT<-abs(DF$SH1-DF$SH2)<1.2
ggplot(data = DF,aes(x=SH2,y=SH1,color=OUT))+geom_point()
```

# Beta diversity correlation

This chunk uses Procrustes analysis to check whether the composition of the community is consistent in the technical replicates. 

```{r}


BC1<-as.matrix(vegdist(otu_table(transform_sample_counts(dupl.1,function(x) x/sum(x)))))
BC2<-as.matrix(vegdist(otu_table(transform_sample_counts(dupl.2,function(x) x/sum(x)))))
dupl.1.DF<-data.frame(sample_data(dupl.1))
dupl.2.DF<-data.frame(sample_data(dupl.2))
dupl.1.DF$SSUMS<-sample_sums(dupl.1)
dupl.2.DF$SSUMS<-sample_sums(dupl.2)

rownames(dupl.1.DF)<-dupl.1.DF$ID_D
rownames(dupl.2.DF)<-dupl.2.DF$ID_D
dupl.2.DF<-dupl.2.DF[match(rownames(dupl.1.DF),rownames(dupl.2.DF) ),]

rownames(BC1)<-colnames(BC1)<-sample_data(dupl.1)$ID_D
rownames(BC2)<-colnames(BC2)<-sample_data(dupl.2)$ID_D

BC2<-BC2[rownames(BC1),rownames(BC1)]

BC1.pcoa<-pcoa(as.dist(BC1))$vectors
BC2.pcoa<-pcoa(as.dist(BC2))$vectors

PROTEST<-protest(BC1.pcoa,BC2.pcoa)

DF.prot<-data.frame(PROTEST$Y[,1:2],PROTEST$X[,1:2],sample_data(dupl.1),
                    resid=resid(PROTEST))

names(DF.prot)[1:4]<-c("X1","X2","Y1","Y2")

ggplot(DF.prot,aes(x=X1,y=X2))+geom_point()+geom_segment(aes(x=X1,y=X2,xend = Y1, yend = Y2),arrow = arrow(length = unit(0.1,"cm")))

```

# Merging technical replicates

Technical duplicates are merged using the merge.duplicates function so that each sample in the resulting database corresponds to a biological sample. ASVs that were not detected in both duplicates are eliminated with the dupl.concensus function.

```{r}
CONSIST<-dupl.concensus(PHYLOSEQ.DUPL,"ID_D")
CONSIST.merged<-merge.duplicates(PHYLOSEQ=CONSIST,NAMES="ID_D")

IN_DUPLICATED<-taxa_names(PHYLOSEQ.NONDUPL)%in%taxa_names(CONSIST.merged)
IN_DUPLICATED<-IN_DUPLICATED*1
FOUR_PRECENT_TAXA<-transform_sample_counts(PHYLOSEQ.NONDUPL,function(x) ifelse((x/sum(x))>0.04,1,0))

FOUR_PRECENT_TAXA_otutab<-otu_table(FOUR_PRECENT_TAXA)
class(FOUR_PRECENT_TAXA_otutab)<-"matrix"

for(i in 1:dim(FOUR_PRECENT_TAXA_otutab)[2]){
  FOUR_PRECENT_TAXA_otutab[,i]<-(FOUR_PRECENT_TAXA_otutab[,i]+IN_DUPLICATED[i])>0
}

OTU_TAB<-otu_table(PHYLOSEQ.NONDUPL)
OTU_TAB.filtered<-OTU_TAB*FOUR_PRECENT_TAXA_otutab

OTU_TAB.filtered<-otu_table(OTU_TAB.filtered,taxa_are_rows = F)
PHYLOSEQ.NONDUPL2<-PHYLOSEQ.NONDUPL
otu_table(PHYLOSEQ.NONDUPL2)<-OTU_TAB.filtered
PHYLOSEQ.NONDUPL2<-prune_taxa(taxa_sums(PHYLOSEQ.NONDUPL2)>0,PHYLOSEQ.NONDUPL)

sample_names(PHYLOSEQ.NONDUPL2)<-sample_data(PHYLOSEQ.NONDUPL)$ID_D
sample_data(PHYLOSEQ.NONDUPL2)$DUPLICATED<-"NO"
sample_data(CONSIST.merged)$DUPLICATED<-"YES"

PHYLOSEQ_dupl_nondulp<-merge_phyloseq(CONSIST.merged,PHYLOSEQ.NONDUPL2)

```

# Taxonomy

Taxonomy is assigned with RDP classifier using Silva v. 138 as a reference. Final phyloseq is saved.
```{r}

SEQS<-as.character(refseq(PHYLOSEQ_dupl_nondulp))

tax80<-assignTaxonomy(SEQS,refFasta = "~/DB/DADA2/silva_nr99_v138_train_set.fa.gz",minBoot = 80,multithread = TRUE,
                      verbose = TRUE)


save(tax80,file = "/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/tax80.R")

tax80<-tax_table(tax80)

PHYLOSEQ_dupl_nondulp<-merge_phyloseq(PHYLOSEQ_dupl_nondulp,tax80)

save(PHYLOSEQ_dupl_nondulp,file = "/media/kreising/DATA/data/RUN_16_3_2019_areny/PHYLOSEQ_PSEUDO/PHYLOSEQ_dupl_nondulp.R")

```



